---
title: scDesigner Server Logic
embed-resources: true
execute:
  message: false
  warning: false
---

Here are some libraries we'll need.

```{r}
library(scDesigner)
library(SummarizedExperiment)
library(glue)
library(stringr)
```

```{r}
FAMILY_MAP <- list(
    Gaussian = "GaussianCopula",
    Bernoulli = "BernoulliCopula",
    NegBin = "NegBinCopula",
    NegBinIRLS = "NegBinIRLSCopula",
    Poisson = "PoissonCopula",
    pNMF = "PositiveNMF",
    ZeroInflatedNegBin = "ZeroInflatedNegBinCopula",
    ZeroInflatedPoisson = "ZeroInflatedPoissonCopula"
)
```

The functions represent how I imagine the user input will map to the scDesigner
code.

```{r}
make_formulas <- function(formulas) {
    formula_strs <- list()

    for (param in names(formulas)) {
        terms <- c()
        variables <- formulas[[param]]

        for (var in variables) {
            if (!is.null(var$transform) && var$transform$type == "spline") {
                terms <- c(terms, glue("bs({var$variable}, df={var$transform$df})"))
            } else {
                terms <- c(terms, var$variable)
            }
        }

        formula_key <- glue("{param}_formula")
        formula_strs[[formula_key]] <- if (length(terms) > 0) {
            glue("~ {str_c(terms, collapse = ' + ')}")
        } else {
            "~ 1"
        }
    }

    formula_strs
}

generate_data <- function(state) {
    class_name <- FAMILY_MAP[[state$family]]
    simulator_class <- get(class_name)
    fmla <- make_formulas(state$formulas)

    simulator <- do.call(simulator_class$new, fmla)
    simulator <- simulator$fit(state$template, max_epochs = 25)

    if (!is.null(state$n_cells)) {
        ix <- sample(1:nrow(state$template), state$n_cells, replace = TRUE)
        new_obs <- colData(state$template[, ix])
    } else {
        new_obs <- colData(state$template)
    }
    new_data <- simulator$sample(as.data.frame(new_obs))
    rownames(new_data) <- rownames(state$template)
    list(parameters = simulator$parameters(), new_data = new_data)
}
```

Here is a version of a negative binomial model where we use cell_type as a predictor and don't have any change as a function of pseudotime.

```{r}
example_sce <- fetch_pancreas()

formulas <- list(
    mean = list(list(variable = "cell_type")),
    dispersion = list(list(variable = "cell_type")),
    copula = list()
)

state <- list(
    template = example_sce,
    family = "NegBin",
    formulas = formulas
)

result <- generate_data(state)
parameters <- result$parameters
adata_sim <- result$new_data
```

Here are the coefficients for each gene's negative binomial model.

```{r}
parameters$marginal$mean[, 1:10]
```

Here's the estimated copula covariance matrix.

```{r}
parameters$copula$Intercept
```

We can let the copula correlations depend on the cell type.

```{r}
formulas$copula <- list(list(variable = "cell_type"))
state$formulas <- formulas
result <- generate_data(state)
parameters <- result$parameters

parameters$copula$`cell_type[T.Beta]`
```

We can allow the means and dispersions to vary smoothly over pseudotime by using
spline transformations in the predictors. The `make_formulas` function
transforms the object below into strings like `~ bs(pseudotime, df=8)`. I'm
breaking that string down into this structure because I think it might map more
easily onto user inputs (e.g., selecting variables from a list populated by the
`obs` field of the anndata object, then clicking whether a spline should be
used). But internally we just need the formula strings -- let me know if there's
a more natural data structure from a UI perspective.

```{r}
formulas <- list(
    mean = list(list(variable = "pseudotime", transform = list(type = "spline", df = 8))),
    dispersion = list(list(variable = "pseudotime", transform = list(type = "spline", df = 3))),
    copula = list(list(variable = "cell_type"))
)

state <- list(
    template = example_sce,
    family = "NegBin",
    formulas = formulas
)

result <- generate_data(state)
```

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
theme_set(theme_classic())

plot_df <- bind_rows(
  as.data.frame(as.matrix(t(assay(example_sce)))) |> mutate(type = "real", pseudotime = example_sce$pseudotime),
  as.data.frame(as.matrix(t(assay(result$new_data)))) |> mutate(type = "sim", pseudotime = result$new_data$pseudotime)
) |>
  pivot_longer(cols = -c(type, pseudotime), names_to = "gene", values_to = "expression")

plot_df |>
  filter(gene %in% rownames(example_sce)[1:8]) |>
  ggplot(aes(pseudotime, log1p(expression), color = type)) +
  geom_point(alpha = 0.3) +
  facet_wrap(~gene, scales = "free_y", nrow = 2)
```

Here is an example where we use a Gaussian on log transformed data.

```{r}
formulas <- list(
    mean = list(list(variable = "pseudotime", transform = list(type = "spline", df = 8))),
    sdev = list(list(variable = "pseudotime", transform = list(type = "spline", df = 3))),
    copula = list(list(variable = "cell_type"))
)

example_sce_transform <- example_sce
example_sce_transform@assays@data$X <- log1p(example_sce_transform@assays@data$X)

state <- list(
    template = example_sce_transform,
    family = "Gaussian",
    formulas = formulas
)

result <- generate_data(state)
result$parameters$marginal$mean[, 1:10]
```
